Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-09-01 manu-draft.qmd
results revision done, simplified current extent and target tables

Subtidal habitats included hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13]. Intertidal habiats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15]. Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below). This 5-foot contour area was considered the most likely to be impacted by sea-level rise based on current estimates. Since 1946, the St. Petersburg tidal gauge (NOAA gauge 872650) has documented a nearly 20 cm increase in mean tidal height to present day. Projections to the year 2100 suggest sea levels can increase from 58 to 259 cm in the region [@tbep0519].

@@

Opportunity areas for restoration and recommended coverage targets for habitat types were identified by integrating multiple datasets available for the region. First, habitat status and historical trends were quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time. Second, historic habitat restoration efforts conducted in the Tampa Bay watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners. The first two steps were used to identify short-term (2030) goals and long-term (2050) targets for coverage (hectares) of native habitats. Finally, remaining opportunities for restoration were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be impacted by sea-level rise. As such, the approach identifies goals and targets and provides spatially explicit information that identifies where restoration practitioners should pursue projects to meet these goals and targets.

@@

All opportunity areas identified as restorable included reservation restorable (in the coastal stratum), existing conservation restorable, and proposed conservation restorable. These layers were further grouped into their restoration potential as discrete habitat types that could be the goal of future restoration projects. The restoration potential was defined using the coastal stratum, soil types, and salinities of adjacent waters.

Compared to vegetation communities, soil characteristics typically change slowly (e.g., decades to centuries) in response to hydrologic impacts, unless physically disturbed [@Osland12; @Stockmann14]. Therefore, soil distributions can be used to estimate historical habitat distribution and restoration potential. @Ries14 created a soils suitability analysis for wetland mitigation and restoration and classified all soils in the Tampa Bay watershed as xeric, mesic, or hydric. The mesic and hydric categories were combined to represent wetland restoration potential, while the xeric category was used to represent upland restoration potential.

A distinction was made between tidal and freshwater wetland restoration potential by intersecting the mesic and hydric soils with the coastal stratum. Mesic or hydric soils that occur below the 5-foot contour were assigned a restoration potential for tidal wetlands, while mesic or hydric soils occurring above the 5-foot contour were assigned a restoration potential for freshwater wetlands. A final restoration potential category was created for low salinity salt marshes (*Juncus* spp.) in the coastal stratum. These included restorable lands with mesic or hydric soils and also lands adjacent to waters with average salinity values less than 18 psu [@Eleuterius79]. The salinity values were identified using salinity isohalines created from a long-term water quality data set provided by the Environmental Protection Commission of Hillsborough County.

Two distinct mapping products were created from the above analysis. The first was an opportunities map that showed areas in the watershed identified as existing conservation native, existing conservation restorable, proposed conservation native, proposed conservation restorable, reservation native, and reservation restorable. The second was a map that identified the restorable lands (either proposed, existing, or reservation) based on their restoration potential as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands (which included salt marshes). All analyses were conducted using the R statistical programming language [@RCT22], specifically leveraging tools from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [@Pebesma18]. The workflows and data are provided in an open-access repository available on GitHub (<https://github.com/tbep-tech/hmpu-workflow>) for use by others [@Beck22].

@@

The map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap). Native lands (existing conservation native, proposed conservation native, reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW).  Similarly, restorable lands (existing conservation restorable, proposed conservation restorable, reservation restorable) totalled `r areas$restorable$areaha` (`r areas$restorable$areaper` of watershed).  Understandably, most of the native and restorable lands occurred in undeveloped areas of the watershed, particularly in the north and southwest. Existing conservation lands (existing conservation native, existing conservation restorable) totalled `r areas$existing$areaha` (`r areas$existing$areaper` of the watershed) and proposed conservation lands (proposed conservation native, proposed conservation restorable) totalled `r areas$proposed$areaha` (`r areas$proposed$areaper` of the watershed).  Reservation areas in the coastal stratum (reservation native, reservation restorable) totalled `r areas$reservation$areaha` (`r areas$reservation$areaper` of the watershed).

Combining the restorable lands in @fig-oppmap with the coastal stratum, soils, and salinity data provided a spatial summary of the restoration potential grouped by habitat type (@fig-restmap).

Table 3

Integration of all of the opportunity layers provides a summary of the available restoration for the TBEP habitats of interest (Table 5). The "Native Habitats" columns show the total current extent as well as the portion of the current extent occurring on existing conservation lands and proposed conservation lands, respectively. The "Restorable Habitats" columns show the total restoration opportunity as well as the portion of the total restoration opportunity on existing and proposed conservation lands. The majority of restoration opportunities on existing conservation lands are for native uplands and freshwater wetlands (Table 5). However, there are approximately 627 ha of emergent tidal wetland restoration opportunities on existing conservation lands, including 530 ha applicable to higher salinity mangrove forests and salt barrens, and 16 ha applicable to lower salinity salt marsh (e.g., *Juncus roemerianus*) restoration and creation.

The best estimates of total restoration opportunities for urban shorelines and tidal tributaries are provided by the Tampa Bay Living Shoreline Suitability Model (LSSM) prepared by the Florida Fish and Wildlife Conservation Commission [@Boland18]. There is 2,566 km of shoreline in Tampa Bay, and approximately 33% is recommended for living shoreline enhancement.

### Establishment of goals and targets

Table 4

Recommended targets were based on habitat status and trends, habitat restoration data, identified restoration opportunities, and current and anticipated trends in development, available funding, and regulations [@Hilderbrand05; @Hobbs07].

The targets and goals (Table 6) identify where the 2017 CCMP Bay Habitat goals and strategies and the 2021-2025 Strategic Plan [@Burke20] thriving habitats and abundant wildlife programmatic priorities can be implemented. However, it also recognizes that the identified habitat protection and restoration areas will change over time, and will be revisited on a 10-year recurring cycle. A 30-year planning horizon (2050) is also identified based upon sea-level rise projections developed specifically for Tampa Bay [@Burke19]. The coastal stratum (from the existing mean low water line to the approximate 5-foot contour) is projected to directly experience the effects of sea-level rise by 2050, and is the primary focus area for coastal habitat protection and restoration activities. Land acquisition or protection (through conservation easements or other mechanisms) will be needed to ensure completion of targets and goals for both salt marsh and upland habitats (Fig 3).

Targets that maintain current coverage ("hold-the line strategies") were identified for habitats that appear to currently be sustained at acceptable levels. Evolving information such as an Oyster Habitat Suitability Index [@Boswell12] and ongoing mapping exercises will be used to identify optimal locations to conduct restoration activities that help achieve targets and goals. Coordination with the establishment of state-mandated minimum flows [@Munson07] will be necessary for restoration and maintenance of low-salinity salt marsh habitats that will experience higher salinities and rapid transition to mangroves under existing sea-level rise scenarios [@Sherwood14; @Geselbracht15].

>>>>

Subtidal habitats included hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13]. Intertidal habiats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15]. Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below). This 5-foot contour area was considered the most likely to be impacted by sea-level rise based on current estimates. Since 1946, the St. Petersburg tidal gauge (NOAA gauge 872650) has documented a nearly 20 cm increase in mean tidal height to present day. Projections to the year 2100 suggest sea levels can increase from 58 to 259 cm in the region [@Burke19].

@@

Opportunity areas for restoration and recommended coverage targets for habitat types were identified by integrating multiple datasets available for the region. First, habitat status and historical trends were quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time. Second, historic habitat restoration efforts conducted in the Tampa Bay watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners. The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for coverage (hectares) of native habitats. Finally, remaining opportunities for restoration were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be impacted by sea-level rise. As such, the approach identifies goals and targets and provides spatially explicit information that identifies where restoration practitioners should pursue projects to meet these goals and targets.

@@

All opportunity areas identified as restorable included reservation restorable (in the coastal stratum), existing conservation restorable, and proposed conservation restorable.  To identify discrete habitat types that could be the goal of future restoration projects, restorable lands in the coastal stratum and on existing conservation areas (reservation restorable and existing conservation restorable) were further grouped into their restoration potential. Proposed conservation areas were excluded from the analysis to provide a more confident assessment of restoration potential in areas that have already been acquired (i.e., existing conservation) or are immediately threatened by sea-level rise and/or coastal development (i.e., the coastal stratum). The restoration potential was defined using soil types, described as follows.

Compared to vegetation communities, soil characteristics typically change slowly (e.g., decades to centuries) in response to hydrologic impacts, unless physically disturbed [@Osland12; @Stockmann14]. Therefore, soil distributions can be used to estimate historical habitat distribution and restoration potential. @Ries14 created a soils suitability analysis for wetland mitigation and restoration and classified all soils in the Tampa Bay watershed as xeric, mesic, or hydric. The mesic and hydric categories were combined to represent wetland restoration potential, while the xeric category was used to represent upland restoration potential. A distinction was made between tidal and freshwater wetland restoration potential by intersecting the mesic and hydric soils with the coastal stratum. Mesic or hydric soils that occur below the 5-foot contour were assigned a restoration potential for tidal wetlands, while mesic or hydric soils occurring above the 5-foot contour were assigned a restoration potential for freshwater wetlands.

Two distinct mapping products were created from the above analysis. The first was an opportunities map that showed areas in the watershed identified as existing conservation native, existing conservation restorable, proposed conservation native, proposed conservation restorable, reservation native, and reservation restorable. The second was a map that identified the restorable lands (either proposed, existing, or reservation) based on their restoration potential as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands (which included salt marshes). All analyses were conducted using the R statistical programming language [@RCT22], specifically leveraging functions from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [@Pebesma18]. The workflows and data are provided in an open-access repository available on GitHub (<https://github.com/tbep-tech/hmpu-workflow>) for use by others [@Beck22].

@@

Summaries of the current extent of each habitat type are shown in Table 3 as representations of the opportunity areas and restoration potential.  The extent of each habitat in existing conservation lands and proposed conservation lands is shown.  Summaries of the restoration potential under existing and proposed conservation lands is also shown. The majority of restoration opportunities on existing conservation lands are for native uplands and freshwater wetlands. Less opportunities exist for intertidal wetlands (mangrove forests, salt barrens, and salt marshes). These summaries are also shown spatially in @fig-oppmap and @fig-restmap.

The map of the remaining opportunity areas provided an areal estimate of where practitioners could target future restoration projects (@fig-oppmap). Native lands (existing conservation native, proposed conservation native, reservation native) totaled `r areas$native$areaha` (`r areas$native$areaper` of the watershed above MLLW).  Similarly, restorable lands (existing conservation restorable, proposed conservation restorable, reservation restorable) totalled `r areas$restorable$areaha` (`r areas$restorable$areaper` of the watershed).  Understandably, most of the native and restorable lands occurred in undeveloped areas in the north and southwest. Existing conservation lands (existing conservation native, existing conservation restorable) totalled `r areas$existing$areaha` (`r areas$existing$areaper` of the watershed) and proposed conservation lands (proposed conservation native, proposed conservation restorable) totalled `r areas$proposed$areaha` (`r areas$proposed$areaper` of the watershed).  Reservation areas in the coastal stratum (reservation native, reservation restorable) totalled `r areas$reservation$areaha` (`r areas$reservation$areaper` of the watershed).

Combining the restorable lands on existing conservation areas and in the coastal stratum with soils data provided a spatial summary of the restoration potential grouped by habitat type (@fig-restmap).  A total of `r areas$potentialall$areaha` (`r areas$potentialall$areaper` of the watershed) were identified as potentially restorable as coastal uplands (`r areas$potential[['Coastal Uplands']]$areaha`, `r areas$potential[['Coastal Uplands']]$areaper` of the watershed), freshwater wetlands (`r areas$potential[['Freshwater Wetlands']]$areaha`, `r areas$potential[['Freshwater Wetlands']]$areaper` of the watershed), native uplands (`r areas$potential[['Native Uplands']]$areaha`, `r areas$potential[['Native Uplands']]$areaper` of the watershed), or tidal wetlands (`r areas$potential[['Tidal Wetlands']]$areaha`, `r areas$potential[['Tidal Wetlands']]$areaper` of the watershed).

### Establishment of targets and goals

Identifying short-term (2030) targets and long-term (2050) goals for the extent of native habitats in Tampa Bay was informed by the assessment of current extents, past trends, and relative effort for past restoration and enhancement projects. Table 4 shows the targets and goals identified through this analysis and the associated rationale. For example, the targets and goals are established based on the current extent and the restoration potential.  If restoration potential exists and coverage restored from past projects suggests a reasonable level of effort, the targets and goals reflect the current extent plus the restoration opportunity.  Conversely, other habitats with no identified restoration opportunity, or current extents are sufficient (e.g., mangrove forests), were assigned targets and goals similar to the current extent, i.e., these habitats should be protected and restoration is not recommended.  Implicit in the targets and the goals is recurring re-assessment over time to evaluate progress and adjust expectations as appropriate.
